---
title: "Prediction Motivation Project"
author: "Kelly O’Brien"
date: "September 26, 2015"
output: html_document
---

This project will analyze the Weight Lifting Exercise Dataset to find a machine learning algorithm to
predict whether a given person is performing an exercise correctly.  Information about the data set can be
found at [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har). 

```{r include=FALSE, cache=FALSE}
# Load the necessary libraries
library(ggplot2);library(caret);library(randomForest)
```

##Load and Clean Data

The first step in the process is to read the comma separated varable (csv) data from the supplied data files.
An inspection of the file contents using a text editor shows that there are a number of problems
with missing data and also invalid data which is marked with the string *#DIV/0!* rather than the
traditional *NA*.

We will avoid loading strings as factors and assign both *NA* and *#DIV/0!* as missing values while loading the
data.  Note that the data files must be in the current directory for the load.

```{r}
#setwd("/Users/kobrien/DataScientist/MachineLearning")
trainingIn <- read.csv(file="pml-training.csv",head=TRUE,sep=",",na.strings = c("NA","#DIV/0!"),stringsAsFactors=FALSE)
# Convert the classe classification variable back to a factor for the model
trainingIn$classe = as.factor(trainingIn$classe) # Fix this, it needs to be a factor

# This data will be used as the final validation of the model.  It does not have the classe column
finalTestingIn <- read.csv(file="pml-testing.csv",head=TRUE,sep=",",na.strings = c("NA","#DIV/0!"),stringsAsFactors=FALSE)
```

There are a number of features that relate only to the collection of the training data (*unrelatedCols* below).
Examples incude the person's name that generated the training data and the date and time when the training data
was colected.  These kinds of features are not applicable for prediction.

There are also a number of features (columns) that are missing most of their values (*naCols* below).  We will
discard any feature that is missing at least half of its data.

The cleaned data is placed into the training and finalTesting variables.

```{r}
# Filter out unrelated information that we don't want to include
unrelatedCols = c("X","user_name","raw_timestamp_part_1","raw_timestamp_part_2","cvtd_timestamp","new_window","num_window")

# Get rid of any feature that has at least half of the values missing
naCols = names(trainingIn)[which(colSums(is.na(trainingIn)) > dim(trainingIn)[1] / 2)]

# Now get just the columns that are not part of the unrelatedCols or the naCols
goodFeatures = names(trainingIn)[which(!names(trainingIn) %in% append(unrelatedCols,naCols))]
training = trainingIn[,goodFeatures]

# We must perform the same operations on the final testing data
goodFeatures = names(finalTestingIn)[which(!names(finalTestingIn) %in% append(unrelatedCols,naCols))]
finalTesting = finalTestingIn[,goodFeatures]
```

We are left with `r dim(training)[2]` columns of the original `r dim(trainingIn)[2]` columns
which is a considerable reduction in complexity.

##Data Exploration

The next step is to determine which of the remaining features should be used as predictors.  We can check
for which features (predictors) are correlated using the cor function.

A correlation between .85 and 1.0 (not including 1.0) gives us the highly correlated predictors.
The output below shows that there are several features that are strongly correlated which means 
the correlated features are providing redundant information.

```{r}
# Get the pred(ictor) cor(relations)
predcor <- cor(training[, -53])
# This will show how many correlations are between .85 and 1 as well as between -.85 and -1
predcor[abs(predcor) > .85 & abs(predcor) < 1.0]
```

Reading through the output of *print(predcor)*, it appears that the feature *roll_belt* shows a
strong correlation with several items.  The *roll_belt* item is not required since 
it brings no information that is not supplied by the correlated items.

```{r}
#head(1,predcor)
predcor["roll_belt","yaw_belt"]
predcor["roll_belt","total_accel_belt"]
predcor["roll_belt","accel_belt_y"]
predcor["roll_belt","accel_belt_z"]
predcor["roll_belt","accel_belt_z"]
```

Similarly, the *total_accel_belt* has a high correlation with the *accel_belt_y* and *accel_belt_z*.  Neither
the *accel_belt_y* or *accel_belt_z* have strong correlations with any other remaining features.

```{r}
predcor["total_accel_belt","accel_belt_y"]
predcor["total_accel_belt","accel_belt_z"]
```

We'll delete the *roll_belt* and *total_accel_belt* features from the data sets.  

```{r}
goodFeatures <- names(training)[which(!names(training) %in% c("total_accel_belt","roll_belt"))]
training <- training[,goodFeatures]

goodFeatures <- names(finalTesting)[which(!names(finalTesting) %in% c("total_accel_belt","roll_belt"))]
finalTesting <- finalTesting[,goodFeatures]
```

##Training

One of the most accurate algorithms is the random forest.  Unfortunately, using the caret
train method with random forest is too slow to be practical since it will attempting to determine the
best tuning parameters by refitting the model multiple times.

Instead, we'll use the *randomForest* method directly which will give us a straightforward implementation
without the extra tuning.  It completes in a fraction of the time that the caret train method requires.

```{r}
# Create the model
modelFit <- randomForest(classe ~ ., data = training, ntree = 2048)
# Show the results (including the out of sample error)
print(modelFit)
```


##Model Results

As can be seen above, the random forest algorithm produces an *out of bag (OOB)* error estimate
of **0.31%** which is very good predicted performance.  No further tuning is necessary and the
selected algorithm is more than adequate for our purposes.

Note that the out of bag error is equivalent to the use of cross validation estimation and this is the
estimate of error for the model.

This is quite an accurate model although it does take quite a while to generate the model with 
the randomForest method.

The final test data can be processed using *predict(modelFit,finalTesting)*.  The results of the
final prediction had an actual error rate of 0% (20/20 correct).

